pipeComp, a general framework for the evaluation of computational pipelines, reveals performant single cell RNA-seq preprocessing tools

### 2020-08-06 bmc_article.tex
Updates from Overleaf

\begin{abstract}

We present \textit{pipeComp} (\url{https://github.com/plger/pipeComp}), a flexible R framework for pipeline comparison handling interactions between analysis steps and relying on multi-level evaluation metrics. We apply it to the benchmark of single-cell RNA-sequencing analysis pipelines using simulated and real datasets with known cell identities, covering common methods of filtering, doublet detection, normalization, feature selection, denoising, dimensionality reduction and clustering. \textit{pipeComp} can easily integrate any other step, tool, or evaluation metric, allowing extensible benchmarks and easy applications to other fields, as we demonstrate through a study of the impact of removal of unwanted variation on differential expression analysis.

@@

Here, we develop a flexible R framework for pipeline comparison and apply it to the evaluation of the various steps of analysis leading from an initial count matrix to a cluster assignment, which are critical in a wide range of applications. We collected real datasets of known cell composition (Table S\ref{tab:table1}) and used a variety of evaluation metrics to investigate in a multilevel fashion the impact of various parameters and variations around a core scRNAseq pipeline. Although we use some datasets based on other protocols, our focus is especially on droplet-based datasets that do not include exogenous control RNA (i.e. spike-ins); see Table S\ref{tab:table1} and Additional File 1: Fig S1 for more details. In addition to previously-used benchmark datasets with true cell labels \cite{duoClustering2018,tianMixology2018}, we simulated two datasets with a hierarchical subpopulation structure based on real 10x human and mouse data using \textit{muscat} \cite{CrowellMuscat2019}. 
Since graph-based clustering \cite{satijaSeurat2015} was previously shown to consistently perform well across several datasets \cite{duoClustering2018,tianMixology2018}, we used the \textit{Seurat} pipeline as the starting point to perform an integrated investigation of: 1) doublet identification; 2) cell filtering; 3) normalization; 4) feature selection; 5) dimension reduction; and, 6) clustering. We compared competing approaches and also explored more fine-grained parameters and variations on common methods. Importantly, the success of methods at a certain analytical step might be dependent on choices at other steps. Therefore, instead of evaluating each step in isolation, we developed a general framework for evaluating nested variations on a pipeline and suggest a multilevel panel of metrics. Finally, we evaluate several recent methods and provide practical recommendations.

@@

The \textit{pipeComp} package defines a pipeline as, minimally, a list of functions executed consecutively on the output of the previous one (Figure \ref{fig:explanation}A; see also Additional File 2). In addition, optional benchmark functions can be set for each step to provide standardized, multi-layered evaluation metrics. Given such a \texttt{PipelineDefinition} object, a set of alternative parameters (which might include different subroutines calling different methods) and benchmark datasets, the \texttt{runPipeline} function then proceeds through all combinations of arguments (or a desired subset), avoiding recomputing the same step twice without the need to save all alternative intermediates, and compiling evaluations (including run time) on the fly (see Additional File 2 for more detail). Variations in a given parameter can be evaluated using all metrics from this point downward in the pipeline. This is especially important because end-point metrics, such as the adjusted Rand index (ARI) \cite{HubertARI1985} for clustering, are not perfect. For example, although the meaning of an ARI score is independent of the number of true subpopulations \cite{steinleyProperties2004}, the number of clusters called is by far the most important determinant of the score: the farther it is from the actual number of subpopulations, the worse the ARI (Additional File 1: Fig S2-3). In this context, one strategy has been to cluster across various resolutions and only consider the results that have the right number of clusters \cite{duoClustering2018}. While this has the virtue of making the results comparable, in practice the number of subpopulations is typically unknown and tools that operate well in this optimal context might not necessarily be best overall. Clustering results are also very sensitive and might not always capture improvements in earlier steps of the pipeline. In addition, we wanted to assess whether the effect of a given parameter alteration is robust to changes in the rest of the pipeline. We therefore monitored several complementary metrics across multiple steps of the process. We proceeded in a step-wise fashion, first testing a large variety of parameters at the early steps of the pipeline along with only a set of mainstream options downstream, then selecting main alternatives and proceeding to a more detailed benchmark of the next step (Figure \ref{fig:explanation}B).

@@

While most methods accurately identified the doublets in the 3 cell lines dataset (mixology10x3cl), the other two datasets proved more difficult (Figure \ref{fig:doublets}A). \textit{scDblFinder} achieved comparable or better accuracy than top alternatives while being the fastest method (Figure \ref{fig:doublets}B). Across datasets, cells called as doublets tended to be classified in the wrong cluster more often than other cells (Figure \ref{fig:doublets}C). We also found that \textit{scDblFinder} improved the accuracy of the clustering across all benchmark datasets even when, by design, the data contained no heterotypic doublet (Figure \ref{fig:filtering}).

@@

Reasoning that the cells we wish to exclude are the cells that would be misclassified, we measured the rate of misclassification of each cell in each dataset across a variety of clustering pipelines, correcting for the median misclassification rate of the subpopulation and then evaluated what properties could be predictive of misclassification (Additional File 1: Fig S8-10). We could not identify any property or simple combination thereof that would be consistently predictive of misclassification; the only feature that stood out across more than one dataset (the Zheng datasets) was that cells with very high read counts have a higher chance of being misclassified{\color{red}, although such a pattern was not observed in the other datasets.}

@@

We next investigated the impact of different normalization strategies. Beside the standard log-normalization included in \textit{Seurat}, we tested \textit{scran}'s pooling-based normalization \cite{lunPooling2016}, \textit{sctransform}'s variance-stabilizing transformation \cite{hafemeisterSCtransform2019}, normalization based on stable genes \cite{linStableGenes2018, deekeStablyExpressed2018}, and \textit{SCnorm} \cite{bacherSCnorm2017} (\textit{scVI}-based normalization \cite{lopezDeep2018} was included separately{\color{red}, as it was not meant for this purpose and follows a slightly different flow}). In addition to log-normalization, the standard \textit{Seurat} clustering pipeline performs per-feature unit-variance scaling so that the PCA is not too strongly dominated by highly-expressed features. We therefore included versions of the different normalization procedures, with or without a subsequent scaling (\textit{sctransform}'s variance-stabilizing transformation involves an approach analogous to scaling). 
\textit{Seurat}'s scaling function (and \textit{sctransform}) also includes the option to regress out the effect of certain covariates. We tested its performance by using the proportion of mitochondrial counts and the number of detected features as covariates. Finally, it has been proposed that the use of stable genes, in particular cytosolic ribosomal genes, can be used to normalize scRNAseq \cite{deekeStablyExpressed2018}. We therefore evaluated a simple linear normalization based on the sum of these genes, as well as nuclear genes.

@@

We further evaluated normalization methods by investigating their impact on the separability of the subpopulations. Since clustering accuracy metrics such as the ARI are very strongly influenced by the number of clusters, we complemented it with silhouette width \cite{RousseeuwSil1987} and mutual information (MI, Figure \ref{fig:norm}B-C). We found most methods (including no normalization at all) to perform fairly well in most of the subpopulations, and that some datasets (e.g. Zhengmix8eq and simMix1) and subpopulations were consistently more challenging across all methods. {\color{red}An exception was \textit{scVI}-based normalization, which performed similarly to other methods on some datasets (Kumar, simMix2, Zhengmix4eq, Zhengmix4uneq), but very poorly on others, in terms of both the subpopulation silhouette and clustering accuracy (Additional File 1: Fig S13; of note, these values were not meant to be used for clustering, and the usage of the \textit{scVI} latent space will be discussed below along with dimensionality reductions)}. Scaling (e.g. `seurat (no scale)' vs `seurat') tended to reduce the average silhouette width of some subpopulations and to increase that of some less distinguishable ones and was generally, but not always, beneficial on the accuracy of the final clustering. Regressing out covariates systematically gave equal or poorer performance on MI, ARI and ARI at true number of clusters, with a strongest negative impact among the Smart-seq datasets (Koh and Kumar). Regressing out covariates using \textit{sctransform} instead, tended to have a much milder effect, possibly owing to its regularization. In general, \textit{sctransform} systematically outperformed other methods and, even though it was developed to be applied to data with unique molecular identifiers (UMI), it also performed fairly well with the Smart-seq protocol (Koh and Kumar datasets).

@@

The impact of the dimensionality reduction methods tested was far greater than that of normalization or feature selection (Additional File 1: Fig S23). Overall, weighting the components by their explained variance tended to increase silhouette width (Additional File 1: Fig S23A) and, in most datasets, clustering accuracy (Additional File 1: Fig S23B). GLM-PCA tended to increase the average silhouette width of already well-defined subpopulations, but \textit{Seurat}'s PCA procedure however proved superior on all metrics. {\color{red} Like GLM-PCA, \textit{scVI}'s Linearly Decoded (LD) data \cite{SvenssonLDVAE2020} and latent space do not explicitly rely on normalized counts. Their performance was however worse than \textit{Seurat}'s PCA in view of the silhouette width (Additional File 1: Fig S15). In terms of clustering accuracy, \textit{scVI}'s latent space performed better that its LD but was nevertheless less accurate than PCA-based alternatives on the datasets with a very low amount of cells (Koh and Kumar) and harboring many subpopulations (simMix1 and Zhengmix8eq).}

@@

Different approaches have been proposed to select the appropriate number of dimensions, from the visual identification of an inflexion point in the curve of the variance explained by each component (`elbow' method), to more complex algorithms. We evaluated the performance of dimensionality estimators implemented in the \textit{intrinsicDimension} package \cite{johnssonDimensionality2015}, as well as common procedures such as the `elbow` method, some scRNAseq-specific methods such as the JackStraw procedure \cite{ChungJackstraw2015} or \textit{scran}'s \texttt{denoisePCA} \cite{LunScran2016}, and the recent application of Fisher Separability analysis \cite{AlberganteSepar2019}. {\color{red} We also included a variation of the molecular cross-validation (MCVR, modified to be applied on normalized data - see methods) approach originally developed to select an calibrate denoising methods \cite{batsonMolecularCrossValidationSingleCell2019}, but which has also been applied to estimate dimensionality \citep{wagnerMonetOpensource2020}.}

@@

{\color{red} \textit{MCVR} and the global maximum likelihood based on translated Poisson Mixture Model (\textit{maxLikGlobal}) tended to produce clusterings with smaller deviations to the true number of subpopulations (Additional File 1: Fig S26). Although most methods performed well on the various clustering measures, \textit{maxLikGlobal} (using 10 or 20 nearest neighbors, although estimates were relatively stable across various values of \textit{k} -- see Additional File 1: Fig S27) provided the dimensionality estimate that best separated challenging subpopulations (Figure \ref{fig:dim}C) and tended to result in the best clustering accuracy (Figure \ref{fig:dim}D). \textit{MCVR} did not perform well for the datasets without UMIs (Koh and Kumar), which violates the assumptions of the method, and while it did increase the separability of some subpopulations (Figure \ref{fig:dim}C), it was detrimental to the classification of some others. We observed a strong negative correlation (\textit{r}=-0.995, p~1.6e-8) between the number of dimensions called by \textit{MCVR} and the log10 average library size of a dataset, which explained the former considerably better than the number of subpopulations.

Of note, the optimal method in our evaluation (\textit{maxLikGlobal})} systematically selected many more components than were associated with the subpopulations in Figure \ref{fig:dim}A, suggesting that although these additional components appear individually uninformative, in combination they nevertheless contribute to classification. An important implication is that the commonly-used elbow method most likely underestimates dimensionality.

@@

The MI score and minimum precision, which are largely independent of the estimated number of clusters, were overall higher for the \textit{walktrap} method (Figure \ref{fig:clust}A, left) over \textit{Seurat}, although when used in combination with some networks it sometimes yielded a lower ARI or minimum precision. The ARI score at the true number of clusters, when available, showed similar performances, especially when using \textit{sctransform}. Because \textit{Seurat}'s resolution parameter had a large impact on the number of clusters identified (Additional File 1: Fig S2 and 24), \textit{Seurat} could always be coerced into producing the right number of clusters. Instead, the number of clusters found by \textit{scran} was considerably less influenced by the available parameters (number of nearest neighbors or steps in the random walk - see Additional File 1: Fig S28), and as a result \textit{scran}-based clustering sometimes never produced a partitioning with the right number of clusters. This observation, along with \textit{scran}'s higher MI score, suggest that \textit{scran} sometimes simply divides a real subpopulation into two clusters (possibly tracking some unknown biological differences) rather than committing misclassification errors. Overall, the \textit{walktrap} method appeared superior to the \textit{fast greedy} algorithm and was generally comparable to \textit{Seurat} clustering, although the latter offered more control over the resolution. A major difference between \textit{walktrap}-based clustering and \textit{Seurat} is the computing time (Figure \ref{fig:clust}B). We found that using the Annoy approximation to the nearest neighbors somewhat reduced computing time at no apparent cost to accuracy. It nevertheless remained considerably slower than \textit{Seurat}. For all methods, some poorly distinguishable subpopulations from both the Zhengmix8eq and simMix1 datasets remained very inaccurately classified in regard to all metrics.

@@

We selected the most important alternatives for each parameter (see Methods) in order to explore their interactions. Unsupervised clustering on the evaluation metrics reveals general patterns in the results as well as the relative importance of the parameters (Additional File 1: Fig S29), with the clustering method having the most important impact on the results. This is however chiefly due to number of clusters called: an ANOVA on all analyses after accounting, in the model, for baseline differences between datasets and the impact of \textit{Seurat}'s resolution parameter, attributed relatively little importance to the clustering method, and instead attributed the largest proportion of variance in the main cluster metrics to the normalization method and the number of dimensions used (Figure \ref{fig:overall}). Part of the impact of the estimated dimensionality depended on its effect on the number of clusters detected, and it indeed explained less variance in ARI at the true number of clusters (Figure \ref{fig:overall}). Nevertheless, the choice of dimensionality (as well as feature selection) came closely behind normalization, suggesting that more attention should be paid to this relatively neglected step.

The greater importance of normalization, dimensionality and feature selection over filtering steps is also visible in the clustering (Additional File 1: Fig S29). The top methods (bottom of Additional File 1: Fig S29) confirm the results of the previous sections, highlighting especially \textit{sctransform}, deviance-based feature selection, and \textit{maxLikGlobal} dimensionality estimates as top methods. Interestingly, while the combination of \textit{sctransform} with \textit{Seurat} clustering was particularly performant for most datasets, for the simulated `simMix1` dataset,  standard \textit{Seurat} normalization combined with \textit{scran}-based clustering proved superior.

We next investigated whether the benefits of some methods were conditional on the choices at other steps. Although analysis of variance identified some significant interactions (Additional File 1: Fig S30), we found that the benefits brought by performant methods were robust to differences in other parts of the pipeline (e.g. Additional File 1: Fig S31). For instance, we confirmed that doublet removal and \textit{sctransform} normalization were beneficial independently of other steps, and that deviance-based feature selection outcompeted \textit{Seurat}'s default selection method independently of changes at other steps (Additional File 1: Fig S31). ANOVA did reveal some significant (negative) interactions (Additional File 1: Fig S30): the top two-way interaction was between \textit{scran} normalization and the elbow method for selection the number of dimensions, and indeed that combination was involved in all of the 30 lowest-ranking pipelines. Furthermore, all significant interactions involved the elbow method, suggesting that some methods were less robust to underestimated dimensionality.

@@

To illustrate the applicability of \textit{pipeComp} to a completely different problem, we benchmarked the impact of surrogate variable analysis (SVA)-based methods on bulk RNAseq differential expression analysis (DEA). The rationale behind such approaches is that the identification of factors determining genome-wide variation (i.e. surrogate variables) and their inclusion in the DEA's generalized linear model might improve DEA, especially in the presence of sources of variation unrelated to the experimental variables. \textit{sva} \citep{leekCapturingHeterogeneityGene2007,leekSvaPackageRemoving2012} and \textit{RUVseq} \citep{rissoNormalizationRNAseq2014a} are two popular packages each implementing different methods to infer these variables. We therefore created a \texttt{PipelineDefinition} with three steps: 1) feature filtering, 2) removal of unwanted variation (i.e. SVA), and 3) DEA using the most established methods  \citep{germainRNAontheBENCHComputationalEmpirical2016}. The package's `pipeComp\_dea' vignette details the creation process. We developed 3 benchmark datasets: 1) the \texttt{ipsc} data is a random selection of 10 vs 10 samples from \citep{carcamoOriveAnalysisTranscriptional2017}, which are very heterogeneous, and foldchanges were manually added to 300 genes; 2) the \texttt{seqc} dataset contains 5 vs 5 samples of two mixtures of the SEQC consortium with different ERCC spike-in mixes \citep{ComprehensiveAssessment2014}, and includes a partially correlated batch effect; 3) A \texttt{simulation} of 8 vs 8 samples with two technical vectors of variation (see Methods for more details). As previously reported \citep{ComprehensiveAssessment2014}, we found that more stringent filtering improved DEA (Additional File 1: Fig S33A-B) and that the different DEA methods had similar performances, especially after using SVA-based methods (Additional File 1: Fig S33B). Including a SVA-based step increased sensitivity across all datasets (Figure \ref{fig:svadea}A and Additional File 1: Fig S33B), particularly in datasets with technical vectors of variation (\textit{seqc} and \textit{simulation}), and benefited all DEA methods in a similar fashion (Figure \ref{fig:svadea}C and Additional File 1: Fig S33C-34A). With the exception of the \texttt{RUVr} method, which led to a high False Discovery Rate (FDR), all other methods showed comparable gains in sensitivity while maintaining a decent error control (if a little above the nominal FDR), even when increasing the number of surrogate variables used (Figure \ref{fig:svadea}B-C). While the FDR did increase in the \textit{ipsc} dataset, we suspect that much of the effect could be due to spurious differences \citep{germainTamingHumanGenetic2017} between the groups which we wrongly assume to be false positives. Nevertheless, the FDR remained acceptable with up to three variables. These results therefore suggest that these methods are robust and can be widely applied.

@@

The set of alternatives found for the overall run (Figure 8 and Additional File 1: Fig S29-31) can be found in the \href{https://github.com/markrobinsonuzh/scRNA_pipelines_paper/blob/master/data/overall_run.Rmd}{paper's repository}. The analysis of variance was performed with the following model:

>>>>

\begin{abstract}
{\color{red} We present \textit{pipeComp} (\url{https://github.com/plger/pipeComp}), a flexible R framework for pipeline comparison handling interactions between analysis steps and relying on multi-level evaluation metrics. We apply it to the benchmark of single-cell RNA-sequencing analysis pipelines using simulated and real datasets with known cell identities, covering common methods of filtering, doublet detection, normalization, feature selection, denoising, dimensionality reduction and clustering. \textit{pipeComp} can easily integrate any other step, tool, or evaluation metric, allowing extensible benchmarks and easy applications to other fields, as we demonstrate through a study of the impact of removal of unwanted variation on differential expression analysis.}

@@

Here, we develop a flexible R framework for pipeline comparison and apply it to the evaluation of the various steps of analysis leading from an initial count matrix to a cluster assignment, which are critical in a wide range of applications. We collected real datasets of known cell composition ({\color{red}Table S\ref{tab:table1}}) and used a variety of evaluation metrics to investigate in a multilevel fashion the impact of various parameters and variations around a core scRNAseq pipeline. Although we use some datasets based on other protocols, our focus is especially on droplet-based datasets that do not include exogenous control RNA (i.e. spike-ins); see {\color{red}Table S\ref{tab:table1}} and Additional File 1: Fig S1 for more details. In addition to previously-used benchmark datasets with true cell labels \cite{duoClustering2018,tianMixology2018}, we simulated two datasets with a hierarchical subpopulation structure based on real 10x human and mouse data using \textit{muscat} \cite{CrowellMuscat2019}. 
Since graph-based clustering \cite{satijaSeurat2015} was previously shown to consistently perform well across several datasets \cite{duoClustering2018,tianMixology2018}, we used the \textit{Seurat} pipeline as the starting point to perform an integrated investigation of: 1) doublet identification; 2) cell filtering; 3) normalization; 4) feature selection; 5) dimension reduction; and, 6) clustering. We compared competing approaches and also explored more fine-grained parameters and variations on common methods. Importantly, the success of methods at a certain analytical step might be dependent on choices at other steps. Therefore, instead of evaluating each step in isolation, we developed a general framework for evaluating nested variations on a pipeline and suggest a multilevel panel of metrics. Finally, we evaluate several recent methods and provide practical recommendations.

@@

The \textit{pipeComp} package defines a pipeline as, minimally, a list of functions executed consecutively on the output of the previous one (Figure \ref{fig:explanation}A; {\color{red}see also Additional File 2}). In addition, optional benchmark functions can be set for each step to provide standardized, multi-layered evaluation metrics. Given such a \texttt{PipelineDefinition} object, a set of alternative parameters (which might include different subroutines calling different methods) and benchmark datasets, the \texttt{runPipeline} function then proceeds through all combinations of arguments (or a desired subset), avoiding recomputing the same step twice without the need to save all alternative intermediates, and compiling evaluations (including run time) on the fly ({\color{red}see Additional File 2 for more details}). Variations in a given parameter can be evaluated using all metrics from this point downward in the pipeline. This is especially important because end-point metrics, such as the adjusted Rand index (ARI) \cite{HubertARI1985} for clustering, are not perfect. For example, although the meaning of an ARI score is independent of the number of true subpopulations \cite{steinleyProperties2004}, the number of clusters called is by far the most important determinant of the score: the farther it is from the actual number of subpopulations, the worse the ARI (Additional File 1: Fig S2-3). In this context, one strategy has been to cluster across various resolutions and only consider the results that have the right number of clusters \cite{duoClustering2018}. While this has the virtue of making the results comparable, in practice the number of subpopulations is typically unknown and tools that operate well in this optimal context might not necessarily be best overall. Clustering results are also very sensitive and might not always capture improvements in earlier steps of the pipeline. In addition, we wanted to assess whether the effect of a given parameter alteration is robust to changes in the rest of the pipeline. We therefore monitored several complementary metrics across multiple steps of the process. We proceeded in a step-wise fashion, first testing a large variety of parameters at the early steps of the pipeline along with only a set of mainstream options downstream, then selecting main alternatives and proceeding to a more detailed benchmark of the next step (Figure \ref{fig:explanation}B).

@@

While most methods accurately identified the doublets in the 3 cell lines dataset (mixology10x3cl), the other two datasets proved more difficult (Figure \ref{fig:doublets}A). \textit{scDblFinder} achieved comparable or better accuracy than top alternatives while being the fastest method (Figure \ref{fig:doublets}B). Across datasets, cells called as doublets tended to be classified in the wrong cluster more often than other cells (Figure \ref{fig:doublets}C). We also found that {\color{red} \textit{scDblFinder} clearly improved the accuracy of the clustering across datasets that are expected to have heterotypic doublets (mixology datasets), and even in a simulation (simMix1) that should not contain any (Figure \ref{fig:filtering}). Across the datasets that, being FACS-sorted, should not contain such doublets, doublet removal tended not to robustly impact clustering accuracy, however in few datasets (ZhengMix4eq and Zhengmix4uneq, and simMix2 with a very small difference) doublet removal performed worse when combined with lenient or no filtering.}

@@

Reasoning that the cells we wish to exclude are the cells that would be misclassified, we measured the rate of misclassification of each cell in each dataset across a variety of clustering pipelines, correcting for the median misclassification rate of the subpopulation and then evaluated what properties could be predictive of misclassification (Additional File 1: Fig S8-10). We could not identify any property or simple combination thereof that would be consistently predictive of misclassification; the only feature that stood out across more than one dataset (the Zheng datasets) was that cells with very high read counts have a higher chance of being misclassified, although such a pattern was not observed in the other datasets.

@@

We next investigated the impact of different normalization strategies. Beside the standard log-normalization included in \textit{Seurat}, we tested \textit{scran}'s pooling-based normalization \cite{lunPooling2016}, \textit{sctransform}'s variance-stabilizing transformation \cite{hafemeisterSCtransform2019}, normalization based on stable genes \cite{linStableGenes2018, deekeStablyExpressed2018}, and \textit{SCnorm} \cite{bacherSCnorm2017} (\textit{scVI}-based normalization \cite{lopezDeep2018} was included separately, as it was not meant for this purpose and follows a slightly different flow). In addition to log-normalization, the standard \textit{Seurat} clustering pipeline performs per-feature unit-variance scaling so that the PCA is not too strongly dominated by highly-expressed features. We therefore included versions of the different normalization procedures, with or without a subsequent scaling (\textit{sctransform}'s variance-stabilizing transformation involves an approach analogous to scaling). 
\textit{Seurat}'s scaling function (and \textit{sctransform}) also includes the option to regress out the effect of certain covariates. We tested its performance by using the proportion of mitochondrial counts and the number of detected features as covariates. Finally, it has been proposed that the use of stable genes, in particular cytosolic ribosomal genes, can be used to normalize scRNAseq \cite{deekeStablyExpressed2018}. We therefore evaluated a simple linear normalization based on the sum of these genes, as well as nuclear genes.

@@

We further evaluated normalization methods by investigating their impact on the separability of the subpopulations. Since clustering accuracy metrics such as the ARI are very strongly influenced by the number of clusters, we complemented it with silhouette width \cite{RousseeuwSil1987} and mutual information (MI, Figure \ref{fig:norm}B-C). We found most methods (including no normalization at all) to perform fairly well in most of the subpopulations, and that some datasets (e.g. Zhengmix8eq and simMix1) and subpopulations were consistently more challenging across all methods. An exception was \textit{scVI}-based normalization, which performed similarly to other methods on some datasets (Kumar, simMix2, Zhengmix4eq, Zhengmix4uneq), but very poorly on others, in terms of both the subpopulation silhouette and clustering accuracy (Additional File 1: Fig S13; of note, these values were not meant to be used for clustering, and the usage of the \textit{scVI} latent space will be discussed below along with dimensionality reductions). Scaling (e.g. `seurat (no scale)' vs `seurat') tended to reduce the average silhouette width of some subpopulations and to increase that of some less distinguishable ones and was generally, but not always, beneficial on the accuracy of the final clustering. Regressing out covariates systematically gave equal or poorer performance on MI, ARI and ARI at true number of clusters, with a strongest negative impact among the Smart-seq datasets (Koh and Kumar). Regressing out covariates using \textit{sctransform} instead, tended to have a much milder effect, possibly owing to its regularization. In general, \textit{sctransform} systematically outperformed other methods and, even though it was developed to be applied to data with unique molecular identifiers (UMI), it also performed fairly well with the Smart-seq protocol (Koh and Kumar datasets).

@@

The impact of the dimensionality reduction methods tested was far greater than that of normalization or feature selection (Additional File 1: Fig S23). Overall, weighting the components by their explained variance tended to increase silhouette width (Additional File 1: Fig S23A) and, in most datasets, clustering accuracy (Additional File 1: Fig S23B). GLM-PCA tended to increase the average silhouette width of already well-defined subpopulations, but \textit{Seurat}'s PCA procedure however proved superior on all metrics. Like GLM-PCA, \textit{scVI}'s Linearly Decoded (LD) data \cite{SvenssonLDVAE2020} and latent space do not explicitly rely on normalized counts. Their performance was however worse than \textit{Seurat}'s PCA in view of the silhouette width (Additional File 1: Fig S15). In terms of clustering accuracy, \textit{scVI}'s latent space performed better that its LD but was nevertheless less accurate than PCA-based alternatives on the datasets with a very low amount of cells (Koh and Kumar) and harboring many subpopulations (simMix1 and Zhengmix8eq).

@@

Different approaches have been proposed to select the appropriate number of dimensions, from the visual identification of an inflexion point in the curve of the variance explained by each component (`elbow' method), to more complex algorithms. We evaluated the performance of dimensionality estimators implemented in the \textit{intrinsicDimension} package \cite{johnssonDimensionality2015}, as well as common procedures such as the `elbow` method, some scRNAseq-specific methods such as the JackStraw procedure \cite{ChungJackstraw2015} or \textit{scran}'s \texttt{denoisePCA} \cite{LunScran2016}, and the recent application of Fisher Separability analysis \cite{AlberganteSepar2019}. We also included a variation of the molecular cross-validation (MCVR, modified to be applied on normalized data - see methods) approach originally developed to select an calibrate denoising methods \cite{batsonMolecularCrossValidationSingleCell2019}, but which has also been applied to estimate dimensionality \citep{wagnerMonetOpensource2020}.

@@

\textit{MCVR} and the global maximum likelihood based on translated Poisson Mixture Model (\textit{maxLikGlobal}) tended to produce clusterings with smaller deviations to the true number of subpopulations (Additional File 1: Fig S26). Although most methods performed well on the various clustering measures, \textit{maxLikGlobal} (using 10 or 20 nearest neighbors, although estimates were relatively stable across various values of \textit{k} -- see Additional File 1: Fig S27) provided the dimensionality estimate that best separated challenging subpopulations (Figure \ref{fig:dim}C) and tended to result in the best clustering accuracy (Figure \ref{fig:dim}D). \textit{MCVR} did not perform well for the datasets without UMIs (Koh and Kumar), which violates the assumptions of the method, and while it did increase the separability of some subpopulations (Figure \ref{fig:dim}C), it was detrimental to the classification of some others. We observed a strong negative correlation (\textit{r}=-0.995, p~1.6e-8) between the number of dimensions called by \textit{MCVR} and the log10 average library size of a dataset, which explained the former considerably better than the number of subpopulations.

Of note, the optimal method in our evaluation (\textit{maxLikGlobal}) systematically selected many more components than were associated with the subpopulations in Figure \ref{fig:dim}A, suggesting that although these additional components appear individually uninformative, in combination they nevertheless contribute to classification. An important implication is that the commonly-used elbow method most likely underestimates dimensionality.

@@

{\color{red} The MI score, which is largely independent of the estimated number of clusters, was overall higher for the \textit{walktrap} method (Figure \ref{fig:clust}A, left) over \textit{Seurat} and the \textit{fast-greedy} method. Depending on what network it was used with, however, it sometimes yielded a lower ARI or minimum precision. The ARI score at the true number of clusters, when available, showed similar performances, especially when using \textit{sctransform}} Because \textit{Seurat}'s resolution parameter had a large impact on the number of clusters identified (Additional File 1: Fig S2 and 24), \textit{Seurat} could always be coerced into producing the right number of clusters. Instead, the number of clusters found by \textit{scran} was considerably less influenced by the available parameters (number of nearest neighbors or steps in the random walk - see Additional File 1: Fig S28), and as a result \textit{scran}-based clustering sometimes never produced a partitioning with the right number of clusters. This observation, along with \textit{scran}'s higher MI score, suggest that \textit{scran} sometimes simply divides a real subpopulation into two clusters (possibly tracking some unknown biological differences) rather than committing misclassification errors. {\color{red} Overall, the \textit{walktrap} method appeared superior to the \textit{fast greedy} algorithm and was generally comparable to \textit{Seurat} clustering, although the latter offered more control over the resolution.} A major difference between \textit{walktrap}-based clustering and \textit{Seurat} is the computing time (Figure \ref{fig:clust}B). We found that using the Annoy approximation to the nearest neighbors somewhat reduced computing time at no apparent cost to accuracy. It nevertheless remained considerably slower than \textit{Seurat}. For all methods, some poorly distinguishable subpopulations from both the Zhengmix8eq and simMix1 datasets remained very inaccurately classified in regard to all metrics.

@@

We selected the most important alternatives for each parameter (see Methods) in order to explore their interactions. Unsupervised clustering on the evaluation metrics reveals general patterns in the results as well as the relative importance of the parameters (Additional File 1: Fig S29), with the clustering method having the most important impact on the results. This is however chiefly due to number of clusters called: an ANOVA on all analyses after accounting, in the model, for baseline differences between datasets and the impact of \textit{Seurat}'s resolution parameter, attributed relatively little importance to the clustering method, and instead attributed the largest proportion of variance in the main cluster metrics to the {\color{red} normalization method and the number of dimensions used (Figure \ref{fig:overall}). Part of the impact of the estimated dimensionality depended on its effect on the number of clusters detected, and it indeed explained less variance in ARI at the true number of clusters (Figure \ref{fig:overall}). Nevertheless, the choice of dimensionality (as well as feature selection) came closely behind normalization, suggesting that more attention should be paid to this relatively neglected step.}

The greater importance of normalization, dimensionality and feature selection over filtering steps is also visible in the clustering ({\color{red}Additional File 1: Fig S29}). The top methods ({\color{red} bottom of Additional File 1: Fig S29}) confirm the results of the previous sections, highlighting especially \textit{sctransform}, deviance-based feature selection, and \textit{maxLikGlobal} dimensionality estimates as top methods. Interestingly, while the combination of \textit{sctransform} with \textit{Seurat} clustering was particularly performant for most datasets, for the simulated `simMix1` dataset,  standard \textit{Seurat} normalization combined with \textit{scran}-based clustering proved superior.

{\color{red}We next investigated whether the benefits of some methods were conditional on the choices at other steps. Although analysis of variance identified some significant interactions (Additional File 1: Fig S30), we found that the benefits brought by performant methods were robust to differences in other parts of the pipeline (e.g. Additional File 1: Fig S31).} For instance, we confirmed that doublet removal and \textit{sctransform} normalization were beneficial independently of other steps, and that deviance-based feature selection outcompeted \textit{Seurat}'s default selection method independently of changes at other steps (Additional File 1: Fig S31). ANOVA did reveal some significant (negative) interactions ({\color{red} Additional File 1: Fig S30}): the top two-way interaction was between \textit{scran} normalization and the elbow method for selection the number of dimensions, and indeed that combination was involved in all of the 30 lowest-ranking pipelines. Furthermore, all significant interactions involved the elbow method, suggesting that some methods were less robust to underestimated dimensionality.

@@

To illustrate the applicability of \textit{pipeComp} to a completely different problem, we benchmarked the impact of surrogate variable analysis (SVA)-based methods on bulk RNAseq differential expression analysis (DEA). The rationale behind such approaches is that the identification of factors determining genome-wide variation (i.e. surrogate variables) and their inclusion in the DEA's generalized linear model might improve DEA, especially in the presence of sources of variation unrelated to the experimental variables. \textit{sva} \citep{leekCapturingHeterogeneityGene2007,leekSvaPackageRemoving2012} and \textit{RUVseq} \citep{rissoNormalizationRNAseq2014a} are two popular packages each implementing different methods to infer these variables. We therefore created a \texttt{PipelineDefinition} with three steps: 1) feature filtering, 2) removal of unwanted variation (i.e. SVA), and 3) DEA using the most established methods  \citep{germainRNAontheBENCHComputationalEmpirical2016}. The package's `pipeComp\_dea' vignette details the creation process. We developed 3 benchmark datasets: 1) the \texttt{ipsc} data is a random selection of 10 vs 10 samples from \citep{carcamoOriveAnalysisTranscriptional2017}, which are very heterogeneous, and foldchanges were manually added to 300 genes; 2) the \texttt{seqc} dataset contains 5 vs 5 samples of two mixtures of the SEQC consortium with different ERCC spike-in mixes \citep{ComprehensiveAssessment2014}, and includes a partially correlated batch effect; 3) A \texttt{simulation} of 8 vs 8 samples with two technical vectors of variation (see Methods for more details). As previously reported \citep{ComprehensiveAssessment2014}, {\color{red} we found that more stringent filtering improved DEA (Additional File 1: Fig S33A-B) and that the different DEA methods had similar performances, especially after using SVA-based methods (Additional File 1: Fig S33B). Including a SVA-based step increased sensitivity across all datasets (Figure \ref{fig:svadea}A and Additional File 1: Fig S33B)}, particularly in datasets with technical vectors of variation (\textit{seqc} and \textit{simulation}), and benefited all DEA methods in a similar fashion (Figure \ref{fig:svadea}C and Additional File 1: Fig S33C-34A). With the exception of the \texttt{RUVr} method, which led to a high False Discovery Rate (FDR), all other methods showed comparable gains in sensitivity while maintaining a decent error control (if a little above the nominal FDR), even when increasing the number of surrogate variables used (Figure \ref{fig:svadea}B-C). While the FDR did increase in the \textit{ipsc} dataset, we suspect that much of the effect could be due to spurious differences \citep{germainTamingHumanGenetic2017} between the groups which we wrongly assume to be false positives. Nevertheless, the FDR remained acceptable with up to three variables. These results therefore suggest that these methods are robust and can be widely applied.

@@

The set of alternatives found for the overall run ({\color{red}Figure 8 and Additional File 1: Fig S29-31}) can be found in the \href{https://github.com/markrobinsonuzh/scRNA_pipelines_paper/blob/master/data/overall_run.Rmd}{paper's repository}. The analysis of variance was performed with the following model:
